---
title: "Extending with MCP and RAG"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending with MCP and RAG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deepagentsr)
```

MCP, retrieval, and UI integrations are optional. The core package remains
small, while helpers make it straightforward to attach external tools and apps.

## MCP tools

`mcp_tools()` accepts already materialized `deep_tool()` objects, which is
useful when another part of your application has already connected to an MCP
server or converted remote capabilities.

```{r materialized-mcp}
remote_echo <- deep_tool(
  function(text) paste("remote echo:", text),
  name = "remote_echo",
  description = "Echo text through a remote-style tool.",
  side_effects = "network",
  returns = "text"
)

tools <- mcp_tools(tools = list(remote_echo))
tools[[1]]$name
tools[[1]]$fun("hello")
```

When `{mcptools}` is installed, `mcp_tools(config = ...)` can read an MCP
configuration and convert returned `ellmer` tool definitions into
`deepagentsr` tools. If `{mcptools}` or a config file is unavailable, the helper
returns an empty list so optional integration code can degrade cleanly.

```{r mcptools-config, eval = FALSE}
agent <- create_deep_agent(
  model = ellmer::chat_openai(model = "gpt-4.1"),
  tools = c(
    list(local_tool),
    mcp_tools(config = "~/.config/mcptools/config.json")
  )
)
```

## RAG tools

Use `rag_tool()` to wrap any R search function with signature `query` and `k`.
The search function can query a data frame, a database, a vector store, or an
external service.

```{r rag-tool}
documents <- data.frame(
  title = c("tools", "approval", "memory"),
  text = c(
    "Tools expose R functions to the model.",
    "Approval pauses side effects before they run.",
    "Memory files are loaded into the supervisor prompt."
  ),
  stringsAsFactors = FALSE
)

search_docs <- function(query, k = 5) {
  hits <- grepl(query, documents$text, ignore.case = TRUE)
  head(documents[hits, , drop = FALSE], k)
}

search_tool <- rag_tool(search_docs)
search_tool$fun("approval", 2)
```

`rag_tools()` returns a list so it can be concatenated with other tool lists.

```{r rag-tools-list}
agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("search_docs", list(query = "Memory", k = 2)),
    assistant_message("Memory files are prompt context.")
  )),
  tools = rag_tools(search_fun = search_docs)
)

agent$invoke("What are memory files?")$text
```

If `{ragnar}` is installed, pass a Ragnar store to `rag_tools(store = store)`.
The resulting `search_docs` tool calls `ragnar::ragnar_retrieve()`.

```{r ragnar, eval = FALSE}
store <- ragnar::ragnar_store_create(embed = my_embedder)

agent <- create_deep_agent(
  model = ellmer::chat_openai(model = "gpt-4.1"),
  tools = rag_tools(store = store, top_k = 5)
)
```

## Shiny chat apps

The package keeps Shiny out of core imports, but ships a small example app
under `inst/examples/shinychat`. It uses `as_shinychat_stream()` as the stable
adapter point between `DeepAgent` event streams and a Shiny chat UI.

```{r shiny-stream}
chat_agent <- create_deep_agent(
  model = fake_chat(list(assistant_message("hello from the agent")))
)

stream <- as_shinychat_stream(chat_agent$stream("hello"))
class(stream)
```

Run the included app from an installed package:

```{r shiny-app, eval = FALSE}
shiny::runApp(system.file("examples", "shinychat", package = "deepagentsr"))
```

The app uses OpenAI through `{ellmer}` when `OPENAI_API_KEY` is set and falls
back to `fake_chat()` otherwise.

## Live OpenAI smoke tests

The repository includes a guarded live test path for OpenAI model smoke tests.
Keep those tests out of normal package checks and opt in only when credentials
and model access are available.

```{r live-tests, eval = FALSE}
readRenviron(".env")
Sys.setenv(RUN_OPENAI_LIVE_TESTS = "true")
devtools::test(filter = "live-openai")
```

The smoke matrix exercises tool-calling through the same `deep_tool()` and
`ellmer` adapter path used by regular agents, including GPT 4.1 and GPT 5
family models when the account has access.
