---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`deepagentsr` provides a small Deep-Agents-style harness for R. A
`DeepAgent` wraps a model, R tools, a virtual filesystem, subagents, memory,
human approval, and execution events behind one entry point:
`create_deep_agent()`.

This vignette uses `fake_chat()` so it can run without a network connection.
Use the same agent configuration with an `ellmer` chat object when you want
live model calls.

## A deterministic first agent

Tools are ordinary R functions wrapped with `deep_tool()`. If you do not
provide an argument schema, `deepagentsr` infers one from the function formals
and converts it to the provider-specific `ellmer` schema when needed.

```{r first-agent}
library(deepagentsr)

calculator <- deep_tool(
  function(x, y) x + y,
  name = "add",
  description = "Add two numbers."
)

agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("write_todos", list(items = list("Call add", "Answer"))),
    assistant_tool_call("add", list(x = 19, y = 23)),
    assistant_message("The result is 42.")
  )),
  tools = list(calculator)
)

result <- agent$invoke("Add 19 and 23.")
result[c("status", "text")]
```

Every run produces structured events. These events are useful for tests,
logging, UI adapters, and JSONL traces.

```{r event-types}
vapply(result$events, function(event) event$type, character(1))
```

The thread state includes turns, todos, pending interrupts, summaries, and
other runtime data.

```{r state}
state <- agent$get_state(result$thread_id)
state$status
state$todos
```

## Use OpenAI through ellmer

For live calls, pass an `ellmer` chat object or a model factory. The package
does not own provider clients directly; it registers tools with `ellmer` and
lets `ellmer` handle OpenAI, Anthropic, Ollama, and other providers.

```{r openai, eval = FALSE}
library(ellmer)
library(deepagentsr)

# Optional for local development when OPENAI_API_KEY is stored in .env.
readRenviron(".env")

calculator <- deep_tool(
  function(x, y) x + y,
  name = "add",
  description = "Add two numbers."
)

agent <- create_deep_agent(
  model = ellmer::chat_openai(model = "gpt-4.1", echo = "none"),
  tools = list(calculator),
  system_prompt = "You are a concise R assistant."
)

agent$invoke("Use the add tool to calculate 19 + 23.")
```

The guarded live test suite has been run against GPT 4.1 and GPT 5 family
models, including `gpt-4.1`, `gpt-4.1-mini`, `gpt-4.1-nano`, `gpt-5`,
`gpt-5-mini`, `gpt-5-nano`, `gpt-5.1`, and `gpt-5.2`, subject to model access
on the configured OpenAI account.

## Streaming and async entry points

`agent$stream()` returns execution events for a run. When `{coro}` is
installed it returns a generator; otherwise it returns a list-like event
stream.

```{r streaming}
stream_agent <- create_deep_agent(
  model = fake_chat(list(assistant_message("streamed response")))
)

stream <- stream_agent$stream("hello")
events <- if (requireNamespace("coro", quietly = TRUE) &&
  inherits(stream, "coro_generator_instance")) {
  coro::collect(stream)
} else {
  stream
}

vapply(events, function(event) event$type, character(1))
```

`agent$stream_async()` is available when `{promises}` is installed and resolves
to the same event stream.

## Runtime context and middleware

Use `context_schema` to require caller-provided runtime fields. Use
`middleware` callbacks to observe every emitted event.

```{r middleware-context}
seen <- character()

context_agent <- create_deep_agent(
  model = fake_chat(list(assistant_message("ok"))),
  context_schema = list(user_id = "character"),
  middleware = list(function(event) {
    seen <<- c(seen, event$type)
  })
)

context_run <- context_agent$invoke(
  "hello",
  context = list(user_id = "user-123")
)

context_run$status
seen
```
