---
title: "Safety and Human Approval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Safety and Human Approval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deepagentsr)
```

Agent tools can read files, write files, call APIs, or trigger other side
effects. `deepagentsr` keeps those capabilities explicit through three layers:

- side-effect metadata on tools;
- filesystem permission rules;
- human-in-the-loop interrupts and resume decisions.

## Tool-level interrupts

Set `interrupt = TRUE` on a tool when every call should pause before the
function body runs.

```{r tool-interrupt}
sent <- new.env(parent = emptyenv())
sent$value <- FALSE

send_email <- deep_tool(
  function(to) {
    sent$value <- to
    "sent"
  },
  name = "send_email",
  description = "Send an email.",
  side_effects = "network",
  interrupt = TRUE
)

email_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("send_email", list(to = "a@example.com")),
    assistant_message("Email sent.")
  )),
  tools = list(send_email)
)

paused <- email_agent$invoke("Send the email.")
paused$status
paused$pending$tool
sent$value
```

The side effect has not happened yet. Resume the same thread with one of the
decision helpers.

```{r tool-interrupt-resume}
resumed <- email_agent$resume(
  paused$thread_id,
  decision_edit(list(to = "b@example.com"))
)

resumed$status
sent$value
```

Use `decision_approve()` to run the original arguments, `decision_edit()` to
replace arguments, `decision_reject()` to tell the model the call was rejected,
or `decision_respond()` to provide a human answer without running the tool.

```{r reject}
reject_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("send_email", list(to = "a@example.com")),
    assistant_message("I did not send the email.")
  )),
  tools = list(send_email)
)

rejected <- reject_agent$invoke("Send another email.")
reject_agent$resume(rejected$thread_id, decision_reject("Do not send email."))$text
```

## Filesystem permissions

`fs_permission()` rules apply to built-in file tools. A rule can allow, deny, or
interrupt operations on matching virtual paths.

```{r permission-interrupt}
write_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call(
      "write_file",
      list(path = "/reports/out.md", content = "draft")
    ),
    assistant_message("Report saved.")
  )),
  permissions = list(
    fs_permission("write", "/reports/out.md", mode = "interrupt")
  )
)

write_paused <- write_agent$invoke("Write the report.")
write_paused$status
write_paused$pending$kind

write_done <- write_agent$resume(write_paused$thread_id, decision_approve())
write_done$status
```

Denied file operations fail immediately.

```{r permission-deny}
blocked_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("read_file", list(path = "/secret.txt"))
  )),
  backend = memory_backend(list("/secret.txt" = "token")),
  permissions = list(
    fs_permission("read", "/secret.txt", mode = "deny")
  )
)

blocked <- try(blocked_agent$invoke("Read the secret."), silent = TRUE)
inherits(blocked, "try-error")
```

## Filesystem backend safety

`filesystem_backend()` maps all virtual paths under one configured host
directory. It blocks path traversal and, by default, common secret-like paths
such as `/.env`.

```{r filesystem-safety}
root <- tempfile("deepagentsr-fs-")
safe_fs <- filesystem_backend(root)
safe_fs$write("/notes/a.txt", "safe")
safe_fs$read("/notes/a.txt")

secret_write <- try(safe_fs$write("/.env", "OPENAI_API_KEY=..."), silent = TRUE)
inherits(secret_write, "try-error")
```

## Local shell tool

Shell execution is not part of the default runtime. `local_dev_shell_tool()` is
an opt-in local development convenience, not a security sandbox. Register it
only for trusted tasks and restrict commands with `allowed_commands`.

```{r shell-tool, eval = FALSE}
rscript <- file.path(R.home("bin"), "Rscript")
execute <- local_dev_shell_tool(
  root_dir = tempdir(),
  allowed_commands = rscript,
  timeout = 10
)

agent <- create_deep_agent(
  model = ellmer::chat_openai(model = "gpt-4.1"),
  tools = list(execute),
  interrupt_on = list(execute = interrupt_config())
)
```

## Auditing decisions

Interrupts, approvals, rejections, file reads, file writes, and tool results
are emitted as events. Export a thread trace when you need an audit record.

```{r audit-events}
trace_path <- tempfile(fileext = ".jsonl")
export_events_jsonl(write_agent, trace_path, thread_id = write_done$thread_id)

events <- read_events_jsonl(trace_path)
vapply(events, function(event) event$type, character(1))
```
