---
title: "Tools and Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools and Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deepagentsr)
```

Tools and files are the main working surface for a deep agent. User tools give
the model controlled access to R functions. Built-in file tools give it a
virtual workspace for notes, drafts, large results, and transcripts.

## Custom R tools

`deep_tool()` stores the R function, model-facing metadata, side-effect
metadata, interrupt policy, and argument schema. Argument schemas can be
explicit, but the package also infers safe generic schemas from function
formals.

```{r custom-tools}
summarise_text <- deep_tool(
  function(text, uppercase = FALSE) {
    out <- paste("Summary:", substr(text, 1, 24))
    if (uppercase) toupper(out) else out
  },
  name = "summarise_text",
  description = "Summarise a short text snippet.",
  side_effects = "none",
  returns = "text"
)

names(summarise_text$arguments)
summarise_text$fun("deepagentsr wraps tools and state", uppercase = TRUE)
```

When the model is an `ellmer` chat object, the agent registers proxy tools that
preserve the original R function formals. Tool calls still flow through the
agent harness, so permissions, interrupts, events, and offloading continue to
work.

## Built-in filesystem tools

Every agent receives built-in tools for todos and virtual files:
`write_todos()`, `read_todos()`, `ls()`, `read_file()`, `write_file()`,
`edit_file()`, `glob()`, and `grep()`.

```{r builtin-files}
backend <- memory_backend()

file_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call(
      "write_file",
      list(path = "/notes/a.txt", content = "alpha\nbeta\ngamma")
    ),
    assistant_tool_call("grep", list(pattern = "beta", path = "/notes")),
    assistant_message("I wrote and searched the note.")
  )),
  backend = backend
)

file_run <- file_agent$invoke("Create a note and search it.")
file_run$text
backend$read("/notes/a.txt")
backend$grep("beta", "/notes")
```

The default backend is in-memory and is ideal for tests, examples, and
ephemeral agent work.

```{r memory-backend}
mem <- memory_backend(list("/docs/readme.md" = "hello world"))
mem$write("/docs/changelog.md", "initial release")
mem$list("/docs")
mem$glob("*.md", "/docs")
```

## Filesystem and durable backends

`filesystem_backend()` maps virtual paths into one configured host directory.
It blocks path traversal and common secret-like paths by default.

```{r filesystem-backend}
root <- tempfile("deepagentsr-fs-")
fs_backend <- filesystem_backend(root)

fs_backend$write("/reports/out.txt", "saved on disk")
fs_backend$read("/reports/out.txt")
```

Use `rds_store_backend()` when you want a small durable virtual filesystem
without exposing a host directory tree to the agent.

```{r rds-store}
store_path <- tempfile(fileext = ".rds")
store <- rds_store_backend(store_path)
store$write("/memory/fact.txt", "R has first-class functions.")

reopened <- rds_store_backend(store_path)
reopened$read("/memory/fact.txt")
```

`composite_backend()` can route path prefixes to different backends. This is
useful when some paths should be durable and others should remain temporary.

```{r composite-backend}
scratch <- memory_backend()
durable <- rds_store_backend(tempfile(fileext = ".rds"))
combined <- composite_backend(
  default = scratch,
  routes = list("/memory" = durable)
)

combined$write("/scratch.txt", "temporary")
combined$write("/memory/profile.txt", "durable")
combined$list("/", recursive = TRUE)
```

## Context offloading and summarization

Large tool results are automatically written to the backend when they exceed
the configured `offload_tokens` threshold. The model receives a short reference
and can inspect the full value later with `read_file()` or `grep()`.

```{r offloading}
large_result <- deep_tool(
  function() paste(rep("large output", 60), collapse = "\n"),
  name = "large_result",
  description = "Return a large result."
)

offload_backend <- memory_backend()
offload_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call("large_result", list()),
    assistant_message("The result was offloaded.")
  )),
  tools = list(large_result),
  backend = offload_backend,
  context_budget = context_budget(offload_tokens = 10)
)

offload_run <- offload_agent$invoke("Call the large tool.")
offload_run$text
offload_backend$list("/internal/offloads", recursive = TRUE)
```

If `max_input_tokens` is set, old turns can be compacted into a summary while
the full transcript is preserved in `/internal/transcripts`.

```{r summarization}
summary_backend <- memory_backend()
summary_agent <- create_deep_agent(
  model = fake_chat(list(
    assistant_tool_call(
      "write_file",
      list(path = "/a.txt", content = paste(rep("alpha", 80), collapse = " "))
    ),
    assistant_tool_call("read_file", list(path = "/a.txt")),
    assistant_message("done")
  )),
  backend = summary_backend,
  context_budget = context_budget(
    max_input_tokens = 80,
    summarize_at = 0.3,
    keep_recent = 0.25
  )
)

summary_run <- summary_agent$invoke("Create and read a large file.")
vapply(summary_run$state$turns, function(turn) turn$role, character(1))
summary_backend$list("/internal/transcripts", recursive = TRUE)
```

## Checkpoints and event traces

`rds_checkpointer()` persists thread state after completed runs and interrupts.

```{r checkpoints}
checkpoint_path <- tempfile(fileext = ".rds")
checkpointer <- rds_checkpointer(checkpoint_path)

checkpoint_agent <- create_deep_agent(
  model = fake_chat(list(assistant_message("checkpointed"))),
  checkpointer = checkpointer
)

checkpoint_run <- checkpoint_agent$invoke("hello")
checkpointer$list_threads()

restored <- create_deep_agent(
  model = fake_chat(list(assistant_message("unused"))),
  checkpointer = checkpointer
)
restored$load_state(checkpoint_run$thread_id)
restored$get_state(checkpoint_run$thread_id)$status
```

Export execution events to JSON Lines for audit logs or downstream analysis.

```{r jsonl-events}
trace_path <- tempfile(fileext = ".jsonl")
export_events_jsonl(checkpoint_agent, trace_path, thread_id = checkpoint_run$thread_id)

trace <- read_events_jsonl(trace_path)
length(trace)
vapply(trace, function(event) event$type, character(1))
```
